Repeated loss of variation in insect ovary morphology highlights the role of development in life-history evolution

### 2021-03-20 manuscript.Rmd
changes to Rmd files during revision process

---
title: Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution
output:
  bookdown::pdf_document2:
  keep_tex: yes
  latex_engine: pdflatex
  #bookdown::html_document2: default
toc: false
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{xr} \externaldocument{manuscript_supplement}
 \usepackage{float}
 \usepackage{array}
 \usepackage{xcolor,colortbl}
 \floatplacement{figure}{H}
---

@@

The number of offspring an organism can produce is a key component of its evolutionary fitness and life-history. Here we perform a test of the hypothesized trade off between the number and size of offspring using thousands of descriptions of the number of egg-producing compartments in the insect ovary (ovarioles), a common proxy for potential offspring number in insects. We find evidence of a negative relationship between egg size and ovariole number when accounting for adult body size. However in contrast to prior claims, we note that this relationship is not generalizable across all insect clades, and we highlight several factors that may have contributed to this size-number trade off being stated as a general rule in previous studies. We reconstruct the evolutionary history of the nurse cell arrangement within the ovariole, and show that the diversification of ovariole number and egg size have both been largely independent of nurse cell presence or position within the ovariole. Instead we show that ovariole number evolution has been shaped by a series of transitions between variable and invariant states, with multiple independent lineages evolving to have almost no variation in ovariole number. We highlight the implications of these invariant lineages on our understanding of the specification of ovariole number during development, as well as the importance of considering developmental processes in theories of life-history evolution.

@@

Offspring number is a fundamental parameter in the study of life-history[@stearns1992evolution]. This number differs widely between organisms[@stearns1992evolution], and its variation is the foundation for several hypotheses about life-history evolution, including the prediction that there is an evolutionary trade off between the number of offspring and their size (e.g. egg size)[@smith1974optimal;@lack1947significance;@stearns1992evolution]. In insects, the number of egg-producing compartments in the ovary, called ovarioles, has been used as a proxy for potential offspring number in the study of life-history[@honvek1993intraspecific;@berrigan1991allometry;@hodin2009she]. However, without an understanding of the phylogenetic distribution of ovariole number, this hypothesized relationship cannot be assessed across insects. Here we tested for the presence of a general trade off between ovariole number and egg size by collecting thousands of records of ovariole number from the published literature, placing them in a phylogenetic context, and comparing them to other datasets of insect reproductive morphology.

@@

Here we compiled `r nrow(raw_ON)` records of ovariole number from across `r ON %>% distinct(order) %>% nrow()` orders, `r ON %>% distinct(family) %>% nrow()` families, and `r ON %>% filter(species != "") %>% distinct(genus,species) %>% nrow()` species of insects. We combined these data with published datasets of egg size[@church2019dataset], fecundity[@Gilbert2007], body size[@rainford2016phylogenetic], and mode of oogenesis[@buning1994insect] to test hypotheses about reproductive evolution. In these analyses we used an existing phylogeny of insects[@church2019insect] to analyze evolutionary patterns in ovariole number, and found that hypotheses about reproductive trait evolution do not hold generally true across insects. Instead we observe that the phylogenetic distribution of ovariole number suggests a model where the developmental mechanisms that govern ovariole number have shifted between variable and invariant states several times over the course of insect evolution. We propose that the developmental mechanisms used to establish ovariole number in well-studied insects such as _Drosophila melanogaster_ are unlikely to regulate ovariole number in all insects.

@@

We searched the published literature for references to insect ovariole number using a predetermined set of 131 search terms, entered into Google Scholar (scholar.google.com) between June and October of 2019. Each search term consisted of an insect taxonomic group and the words "ovariole number". This list was created to include all insect orders, many large insect families, and groups well-represented in the insect egg dataset[@church2019dataset]. For each search term, we evaluated all publications in the first page of results (ten publications). For 61 search terms that had a large number of informative hits, significant representation in the egg dataset, or that corresponded to very speciose groups, we evaluated an additional 20 publications. The list of search terms is available in the supplementary file 'ovariole_number_search_terms.tsv'.

@@

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and dry adult body mass[@Gilbert2007], average adult body length per insect family[@rainford2016phylogenetic], several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low], and the mode of oogenesis[@buning1994insect]. See supplementary methods section \ref{sec:combining} for details.

@@

The analyses in this manuscript were performed using the insect phylogeny published in Church et al., 2019[@church2019insect], unless otherwise specified. Analyses of insect family-level ovariole number, egg size, and body size were performed using the insect phylogeny published in Rainford et. al, 2014[@rainford2014phylogenetic]. Analyses of Drosophilidae ovariole number, egg size, and body size were performed using a phylogeny newly assembled for this study. See supplementary methods section \ref{sec:trees} for details.

To evaluate the robustness of our results to uncertainty in the phylogenetic relationships, all Phylogenetic Generalized Least Squares (PGLS) analyses were performed 1000 times over a posterior distribution of trees, using a Brownian Motion based covariance matrix in the R package ape (version `r packageVersion('ape')`)[@paradis2004ape] and nlme (version `r packageVersion('nlme')`)[@pinheiro2014r]. For regressions at the species and genus level, we reshuffled and matched records for each iteration to account for variation across records for the same taxon. For regressions at the family level we recalculated the average ovariole number per insect family, randomly downsampling the representation for each family by half. To weight traits by body size, we calculated the phylogenetic residuals[@revell2010phylogenetic] of each trait to body size, and then compared the evolution of these residuals using a PGLS regression. See supplementary methods section \ref{sec:PGLS} for details.

For two regressions comparing egg size to ovariole number while accounting for adult body size, we tested alternative hypotheses of evolution by simulating new data. We considered two such hypotheses: no evolutionary correlation with ovariole number, and a strong correlation with ovariole number (slope of -1). For each trait we simulated 1,000 datasets using evolutionary parameters fit under a Brownian Motion model in the R packages geiger (version `r packageVersion('geiger')`)[@revell2010phylogenetic], and phylolm (version `r packageVersion('phylolm')`)[@tung2014linear].

@@

All phylogenetic regressions were performed using both the maximum clade credibility (MCC) tree. We considered a relationship significant when the MCC p-value was below the threshold 0.01. To assess the robustness of results to uncertainty in phylogenetic relationships, we also report the number of regressions over the posterior distribution of phylogenetic trees that gave a significant result (see Table \ref{tab:PGLS-posterior-egg-ON}).

For two comparisons, we validated that our tests had sufficient statistical power using the selected threshold by comparing the distribution of p-values from regressions of observed data to regressions of data simulated under alternative hypotheses. We compared the results of analyses of our observed to those based on simulated data to evaluate the likelihood of false positives (comparing to data simulatd under no correlation) and false negatives (comparing to data simulated with strong correlation).

@@

Ovariole number varies by at least four orders of magnitude across insects (Fig. 1b). We identify seven insect families with species that have been reported to have more than 1,000 total ovarioles, including several eusocial insects (e.g. queens of the termite species _Hypotermes obscuriceps_, Blattodea: Termitidae[@bugnion1912anatomie], and several ant species, Hymenoptera: Formicidae)[@robertson1995sperm;@schneirla1957comparison] and non-eusocial insects (e.g. the blister beetle _Meloe proscarabaeus_, Coleoptera: Meloidae)[@buning1979trophic]. We also find two independent lineages that have evolved to have only one functional ovariole: dung beetles in the tribe Scarabaeinae (Coleoptera: Scarabaeidae)[@richter1974ovariole], and grass flies in the genus _Pachylophus_ (Diptera: Chloropidae)[@meier1999ovoviviparity;@pollock1996viviparous]. In these insects one of the two ovaries presumably established during embryogenesis is reported to atrophy during development[@pollock1996viviparous;@pluot1979evolution], resulting in an asymmetric adult reproductive system.

@@

Ovariole number has been hypothesized to be negatively correlated with egg size[@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. This hypothesis is based on the predictions that (1) female reproduction is resource-limited, therefore egg size should trade off with egg number, and (2) ovariole number can serve as a proxy for egg number[@montague1981reproductive;@smith1974optimal]. We did not observe a significant negative relationship when comparing egg size and ovariole number across insect species (Fig. 2a, Table \ref{tab:PGLS-posterior-egg-ON}, p-value `r round(pgls_results_egg_ON_name$tTable[[8]],3)`, n=`r pgls_results_egg_ON_name$dims$N`). We also compared egg size and ovariole number across insect genera and did not observe a significant relationship (Fig. \ref{fig:PGLS-egg-ON-genus}, p-value `r round(pgls_results_egg_ON_genus$tTable[[8]],3)`, n=`r pgls_results_egg_ON_genus$dims$N`).

Given that this predicted relationship is often conditioned on taking into account adult body size[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with several measurements of body size, and applied a size correction that accounts for phylogenetic relationships[@revell2009size]. When accounting for adult body mass, we observed a significant negative relationship between egg size and ovariole number (Fig. 2b, \ref{fig:PGLS-Gilbert-name}, p-value `r round(pgls_results_gilbert_genus$tTable[[8]],3)`, n=`r pgls_results_gilbert_genus$dims$N`). To evaluate the robustness of this result, we repeated the analysis 1000 times, taking into account uncertainty in both the phylogeny and trait measurements. Out of 1000 regressions, `r sum(sapply(allometry_gilbert_genus,function(x){x$tTable[[8]]}) < 0.01)` indicated a significant negative relationship (Table \ref{tab:PGLS-posterior-egg-ON}). We performed the same comparison accounting for adult body length, and likewise observed a significant negative relationship (Fig. \ref{fig:PGLS-family-length}, p-value `r gsub(0,"<0.001",round(pgls_results_body_egg_ON_family$tTable[[8]],3))`, n=`r pgls_results_body_egg_ON_family$dims$N`), supported by `r sum(sapply(allometry_body_egg_ON_family,function(x){x$tTable[[8]]}) < 0.01)` of 1000 repeated analyses (Table \ref{tab:PGLS-posterior-egg-ON}).

We further explored these results using two methods: First, to evaluate our findings against alternative evolutionary hypotheses, we compared these results to regressions based on simulated data. Our results showed that when considering body size, the slope of the regression of egg size and ovariole number is lower than we would expect to observe by chance, as assessed by comparing to data simulated with no evolutionary correlation (Fig. \ref{fig:simulated}). However, for both adult body length and dry mass, the slope of the regressions on observed data are not within the range that would be expected under an strong negative correlation (slope of -1 in log-log space, Fig. \ref{fig:simulated}). This suggests the presence of a weak evolutionary relationship bewteeen ovariole number and egg size, when accounting for body size.

Second, we assessed the relationship between egg size and ovariole number, accounting for body size, within four subclades of insects. We found that across Drosophilidae fly species, egg size is indeed strongly negatively correlated with ovariole number when accounting for body size (Fig. 2c, Table \ref{tab:PGLS-posterior-clade}, p-value `r gsub(0,"<0.001",round(pgls_results_DIP$tTable[[8]],3))`, n=`r pgls_results_DIP$dims$N`). For grasshoppers and crickets (Orthoptera), beetles (Coleoptera), and wasps (Hymenoptera), we observed no significant relationship between ovariole number and egg size, even when accounting for body size (Fig. 2d, Table \ref{tab:PGLS-posterior-clade}, \ref{fig:PGLS-Coleoptera-Hymenoptera}, Orthoptera: p-value `r round(pgls_results_ORT$tTable[[8]],3)`, n=`r pgls_results_ORT$dims$N`, Coleoptera: p-value `r round(pgls_results_COL1$tTable[[8]],3)`, n=`r pgls_results_COL1$dims$N`, Hymenoptera: p-value `r round(pgls_results_HYM$tTable[[8]],3)`, n=`r round(pgls_results_HYM$dims$N)`). This indicates that, while a strong negative correlation between egg size and ovariole number exists for some insects, it does not represent a universal pattern across insect clades.

Finally, we tested whether ovariole number is positively correlated with adult body size, and in contrast to previous studies[@honvek1993intraspecific], we found no correlation between ovariole number and adult body mass or length across insects (Fig. \ref{fig:PGLS-body-plot}, Table \ref{tab:PGLS-posterior-body}, body mass: p-value `r round(pgls_results_body_gilbert_name$tTable[[8]],3)`, n=`r pgls_results_body_gilbert_genus$dims$N`, body length: p-value `r round(pgls_results_body_ON_family$tTable[[8]],3)`, n=`r pgls_results_body_ON_family$dims$N`). Of the four subclades considered, only insects in the order Orthoptera had a positive relationship between body size and ovariole number (Table \ref{tab:PGLS-posterior-body}, p-value `r round(pgls_results_body_ORT$tTable[[8]],3)`, n=`r pgls_results_body_ORT$dims$N`).

![**Tests of the hypothesized trade off between egg size and ovariole number.** a, Egg volume (mm^3^) and ovariole number, both log~10~ scale; points represent insect species. See section _Modeling ovariole number evolution_ for discussion of the enrichment of certain low values of ovariole number (points appearing vertically arranged) b, Egg volume and ovariole number, residuals to dry adult body mass, points represent genera. c, Drosophilidae egg volume and ovariole number, residuals to thorax length, points represent species. d, Orthoptera egg volume and ovariole number, residuals to body length, points represent genera.](manuscript_figures/Fig_Egg_V2-01.png)

@@

If the hypothesized trade off between the number and size of offspring is true for insects, then one explanation for the lack of a consistent negative relationship between ovariole number and egg size is that ovariole number may not be a reasonable proxy for offspring number. Previous research has shown that, across individuals within the same species, ovariole number is correlated with certain measurements of fecundity, such as maximum daily rate of egg production for _Drosophila_ [@david1970nombre;@bouletreau1982reproductive], but not others, such as lifetime fecundity[@schmidt2005geographic] or fitness in competition assays[@wayne1997quantitative]. Few studies have compared fecundity and ovariole number across species[@stewart1991reproductive], likely due to the difficulties of measuring fecundity consistently across insects, many of which lay eggs singly and continuously rather than in distinct clutches.

Using a previously reported dataset of lifetime fecundity measurements across insects[@Gilbert2007], we assessed the relationship between lifetime fecundity and ovariole number. We observed a significant positive relationship (Fig. 3, p-value `r round(pgls_results_fecundity_genus$tTable[[8]],3)`, n=`r pgls_results_fecundity_genus$dims$N`), however, a substantial fraction of repeat analyses show these results are not robust to uncertainty (`r sum(sapply(allometry_fecundity_genus,function(x){x$tTable[[8]]}) > 0.01)` of 1000 regressions are not significant, Table \ref{tab:PGLS-posterior-fecundity}). We note that this relationship is largely defined by the absence of insects with high ovariole number and low fecundity (Fig. 3, empty bottom right corner), while for insects with low ovariole number, fecundity varied over more than three orders of magnitude. We interpret our results, in conjunction with those previously reported, to suggest that ovariole number, when variable across insects in a lineage, may be one factor among many influencing the number of eggs produced. However, we caution against using ovariole number as a direct mathematical proxy for offspring number.

@@

Using the dataset compiled here and a previously published phylogeny of insects (Fig. 5a)[@church2019insect], we modeled the rate of evolutionary change in ovariole number (Figs. \ref{fig:arbutus}, \ref{fig:BAMM-analysis-convergence}, \ref{fig:BAMM-analysis-burnin}, \ref{fig:BAMM-analysis-results}). We observed substantial rate heterogeneity in the evolution of ovariole number (Fig. \ref{fig:BAMM-analysis-results}), meaning that for some lineages ovariole number has evolved rapidly where in others, ovariole number has evolved very slowly or not at all. The most striking example of this are the multiple lineages which have independently evolved invariant or near-invariant ovariole number across taxa (e.g. nearly all Lepidoptera have exactly eight ovarioles, Fig. 5b, cyan), from an ancestral variable state. These invariant lineages were identified by finding regions of the phylogeny that experience extremely low rates of ovariole number diversification (Figs. \ref{fig:BAMM-analysis-results}, \ref{fig:BAMM-anlaysis-mean-rates}). Using this approach, we found that invariant ovariole numbers have evolved at least nine times independently across insects, with several subsequent reversals from invariant to variable states (Fig. 5a).

@@

Additionally, the distribution of ovariole numbers across insects is enriched for even numbers of total ovarioles (Fig. 5c). While many insects show asymmetries in the number of ovarioles between the left and right ovaries, all of the invariant lineages are symmetric (at 4, 6, 8, 10, 12, and 14 total ovarioles). Therefore, invariant lineages have near-zero variation when comparing between species, between individuals within a species, and between the left and right ovary within an individual.

>>>>

---
title: Repeated loss of variation in insect ovary morphology highlights the role of development in life-history evolution
output:
  bookdown::pdf_document2:
  #bookdown::word_document2: default
  keep_tex: yes
  latex_engine: pdflatex
  #bookdown::html_document2: default
toc: false
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{xr} \externaldocument{manuscript_supplement}
 \usepackage{float}
 \usepackage{array}
 \usepackage{xcolor,colortbl}
 \floatplacement{figure}{H}
---

@@

The number of offspring an organism can produce is a key component of its evolutionary fitness and life-history. Here we perform a test of the hypothesized trade-off between the number and size of offspring using thousands of descriptions of the number of egg-producing compartments in the insect ovary (ovarioles), a common proxy for potential offspring number in insects. We find evidence of a negative relationship between egg size and ovariole number when accounting for adult body size. However in contrast to prior claims, we note that this relationship is not generalizable across all insect clades, and we highlight several factors that may have contributed to this size-number trade-off being stated as a general rule in previous studies. We reconstruct the evolutionary history of the nurse cell arrangement within the ovariole, and show that the diversification of ovariole number and egg size have both been largely independent of nurse cell presence or position within the ovariole. Instead we show that ovariole number evolution has been shaped by a series of transitions between variable and invariant states, with multiple independent lineages evolving to have almost no variation in ovariole number. We highlight the implications of these invariant lineages on our understanding of the specification of ovariole number during development, as well as the importance of considering developmental processes in theories of life-history evolution.

@@

Offspring number is a fundamental parameter in the study of life-history[@stearns1992evolution]. This number differs widely between organisms[@stearns1992evolution], and its variation is the foundation for several hypotheses about life-history evolution, including the prediction that there is an evolutionary trade-off between the number of offspring and their size (e.g. egg size)[@smith1974optimal;@lack1947significance;@stearns1992evolution]. In insects, the number of egg-producing compartments in the ovary, called ovarioles, has been used as a proxy for potential offspring number in the study of life-history[@honvek1993intraspecific;@berrigan1991allometry;@hodin2009she]. However, without an understanding of the phylogenetic distribution of ovariole number, this hypothesized relationship cannot be assessed across insects. Here we tested for the presence of a general trade-off between ovariole number and egg size by collecting thousands of records of ovariole number from the published literature, placing them in a phylogenetic context, and comparing them to other datasets of insect reproductive morphology.

@@

Here we compiled `r nrow(raw_ON)` records of ovariole number from across `r ON %>% distinct(order) %>% nrow()` orders, `r ON %>% distinct(family) %>% nrow()` families, and `r ON %>% filter(species != "") %>% distinct(genus,species) %>% nrow()` species of insects. We combined these data with published datasets of egg size[@church2019dataset], fecundity[@Gilbert2007;@gilbert2010parental], and body size[@rainford2016phylogenetic], to test hypotheses about \textcolor{red}{the evolutionary trade-off between offspring size and number}. In these analyses we used an existing phylogeny of insects[@church2019insect] to analyze evolutionary patterns in ovariole number, and found that hypotheses about \textcolor{red}{life-history} evolution do not hold generally true across insects. \textcolor{red}{We then combined these data with published observations of the mode of oogenesis}[@buning1994insect]\textcolor{red}{ and reconstructed the evolutionary history of the presence and position of nurse cells that contribute to the oocyte during oogenesis. We tested whether patterns in the distribution of ovariole number, egg size, or egg shape were driven by the evolution of nurse cells, and found no significant results.} Instead we observe that the phylogenetic distribution of ovariole number suggests a model where the developmental mechanisms that govern ovariole number have shifted between variable and invariant states several times over the course of insect evolution. Based on this finding, we propose that the developmental mechanisms used to establish ovariole number in well-studied insects such as _Drosophila melanogaster_ are unlikely to regulate ovariole number in all insects.

@@

We searched the published literature for references to insect ovariole number using a predetermined set of 131 search terms, entered into Google Scholar (scholar.google.com) between June and October of 2019. Each search term was \textcolor{red}{comprised of} an insect taxonomic group and the words "ovariole number". \textcolor{red}{The taxonomic groups used to search included} all insect orders, many large insect families, and taxonomic groups that are well-represented in the insect egg dataset[@church2019dataset]. \textcolor{red}{For each Google Scholar search}, we evaluated all publications in the first page of results (ten publications). For 61 search terms that had a large number of informative hits, significant representation in the egg dataset, or that corresponded to very speciose groups, we evaluated an additional 20 publications. The list of search terms is available in the supplementary file 'ovariole_number_search_terms.tsv'.

@@

\textcolor{red}{For records of ovariole number that reported intraspecific variation in ovariole number, we calculated the percent difference as follows: if ovariole number was reported as a range, percent difference was calculated as the $100 * ((max - min) / median)$; if ovariole number was reported as an average with deviations, percent difference was calculated as $100 * ((2 * deviation) / mean)$. When independent observations of ovariole number for a given species were available across multiple published records, we calculated the percent difference as the $100 * ((max - min) / median)$.}

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and dry adult body mass[@Gilbert2007;@gilbert2010parental;@gilbert2011insect], average adult body length per insect family[@rainford2016phylogenetic], several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low], and the mode of oogenesis[@buning1994insect]. See supplementary methods section \ref{sec:combining} for details.

@@

The analyses in this manuscript were performed using the insect phylogeny published in Church et al., 2019[@church2019insect], unless otherwise specified. \textcolor{red}{For regressions involving body size data that were reported as insect family-level averages,} we used the insect phylogeny published in Rainford et. al, 2014[@rainford2014phylogenetic]. Analyses of Drosophilidae ovariole number, egg size, and body size were performed using a phylogeny newly assembled for this study. See supplementary methods section \ref{sec:trees} for details.

To evaluate the robustness of our results to uncertainty in the phylogenetic relationships, all Phylogenetic Generalized Least Squares (PGLS) analyses were performed 1000 times over a posterior distribution of trees, using a Brownian Motion based covariance matrix in the R package ape (version `r packageVersion('ape')`)[@paradis2004ape] and nlme (version `r packageVersion('nlme')`)[@pinheiro2014r]. For regressions at the species and genus level, we reshuffled and matched records for each iteration to account for variation across records for the same taxon. For regressions at the family level we recalculated the average ovariole number per insect family, randomly downsampling the representation for each family by half. To weight traits by body size, we calculated the phylogenetic residuals[@revell2009size] of each trait to body size, and then compared the evolution of these residuals using a PGLS regression. See supplementary methods section \ref{sec:PGLS} for details.

For two regressions comparing egg size to ovariole number while accounting for adult body size, we tested alternative hypotheses of evolution by simulating new data. We considered two such hypotheses: no evolutionary correlation with ovariole number, and a strong correlation with ovariole number (slope of -1). For each trait we simulated 1,000 datasets using evolutionary parameters fit under a Brownian Motion model in the R packages geiger (version `r packageVersion('geiger')`)[@harmon2007geiger], and phylolm (version `r packageVersion('phylolm')`)[@tung2014linear].

@@

All phylogenetic regressions were performed using the maximum clade credibility (MCC) tree \textcolor{red}{(the tree with highest credibility score from the posterior distribution of the Bayesian analysis)}. We considered a relationship significant when the p-value was below the threshold 0.01. To assess the robustness of results to uncertainty in phylogenetic relationships, we also \textcolor{red}{repeated these analyses over the posterior distribution of phylogenetic trees} and report the number of regressions that gave a significant result (see Table \ref{tab:PGLS-posterior-egg-ON}).

For two comparisons, we validated that our tests had sufficient statistical power using the selected threshold by comparing the distribution of p-values from regressions of observed data to regressions of data simulated under alternative hypotheses. We compared the results of analyses of our observed to those based on simulated data to evaluate the likelihood of false positives (comparing to data \textcolor{red}{simulated} under no correlation) and false negatives (comparing to data simulated with strong correlation).

@@

Ovariole number varies by at least four orders of magnitude across \textcolor{red}{insect species} (Fig. 1b). We identify seven insect families with species that have been reported to have more than 1,000 total ovarioles, including several eusocial insects (e.g. queens of the termite species _Hypotermes obscuriceps_, Blattodea: Termitidae[@bugnion1912anatomie], and several ant species, Hymenoptera: Formicidae)[@robertson1995sperm;@schneirla1957comparison] and non-eusocial insects (e.g. the blister beetle _Meloe proscarabaeus_, Coleoptera: Meloidae)[@buning1979trophic]. We also find two independent lineages that have evolved to have only one functional ovariole: dung beetles in the tribe Scarabaeinae (Coleoptera: Scarabaeidae)[@richter1974ovariole], and grass flies in the genus _Pachylophus_ (Diptera: Chloropidae)[@meier1999ovoviviparity;@pollock1996viviparous]. In these insects one of the two ovaries presumably established during embryogenesis is reported to atrophy during development[@pollock1996viviparous;@pluot1979evolution], resulting in an asymmetric adult reproductive system. \textcolor{red}{We also evaluated intraspecific variation in ovariole number, and found that, for species for which it has been reported, the average percent difference number within species is between 10\% and 100\% of the median value (Fig. \ref{fig:intraspecific-variation}).}

@@

Ovariole number has been hypothesized to be negatively correlated with egg size[@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. This hypothesis is based on the predictions that (1) female reproduction is resource-limited, therefore egg size should trade off with egg number, and (2) ovariole number can serve as a proxy for egg number[@montague1981reproductive;@smith1974optimal]. We did not observe a significant negative relationship when comparing egg size and ovariole number across insect species (Fig. 2a, Table \ref{tab:PGLS-posterior-egg-ON}, p-value `r round(pgls_results_egg_ON_name$tTable[[8]],3)`, n=`r pgls_results_egg_ON_name$dims$N`). We also compared egg size and ovariole number, \textcolor{red}{combining data from species within the same genus to increase sample size,} and again did not observe a significant relationship (Fig. \ref{fig:PGLS-egg-ON-genus}, p-value `r round(pgls_results_egg_ON_genus$tTable[[8]],3)`, n=`r pgls_results_egg_ON_genus$dims$N`). \textcolor{red}{To verify this finding was not driven by the high ovariole numbers seen in the queens of some eusocial insects, we repeated this comparison excluding insects from families with eusocial representatives, with the same result (Fig. \ref{fig:PGLS-egg-ON-genus-noneusocial}, p-value `r round(pgls_results_egg_ON_genus_noneusocial$tTable[[8]],3)`, n=`r pgls_results_egg_ON_genus_noneusocial$dims$N`).}

Given that this predicted relationship is often conditioned on body size, \textcolor{red}{which is predicted to limit total potential reproductive investment}[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with data on insect adult body mass[@Gilbert2007;@gilbert2010parental;@gilbert2011insect] and length[@rainford2016phylogenetic]. When accounting for adult body mass, we observed a significant negative relationship between egg size and ovariole number (Fig. 2b, \ref{fig:PGLS-Gilbert-name}, p-value `r round(pgls_results_gilbert_genus$tTable[[8]],3)`, slope `r round(pgls_results_gilbert_genus$coefficients[[2]],3)`, n=`r pgls_results_gilbert_genus$dims$N`). To evaluate the robustness of this result, we repeated the analysis 1000 times, taking into account uncertainty in both the phylogeny and trait measurements. Out of 1000 regressions, `r sum(sapply(allometry_gilbert_genus,function(x){x$tTable[[8]]}) < 0.01)` indicated a significant negative relationship (Table \ref{tab:PGLS-posterior-egg-ON}). We performed the same comparison accounting for adult body length, and likewise observed a significant negative relationship (Fig. \ref{fig:PGLS-family-length}, p-value `r gsub(0,"<0.001",round(pgls_results_body_egg_ON_family$tTable[[8]],3))`, slope `r round(pgls_results_body_egg_ON_family$coefficients[[2]],3)`, n=`r pgls_results_body_egg_ON_family$dims$N`), supported by `r sum(sapply(allometry_body_egg_ON_family,function(x){x$tTable[[8]]}) < 0.01)` of 1000 repeated analyses (Table \ref{tab:PGLS-posterior-egg-ON}).

We further explored these results using two methods: First, to evaluate our findings against alternative evolutionary hypotheses, we compared these results to regressions based on simulated data. Our results showed that when considering body size, the slope of the regression of egg size and ovariole number is \textcolor{red}{more negative} than we would expect to observe by chance, as assessed by comparing to data simulated with no evolutionary correlation (Fig. \ref{fig:simulated}). However, for both adult body length and dry mass, the slope of the regressions on observed data are not within the range that would be expected under a strong negative correlation (slope of -1 in log-log space, Fig. \ref{fig:simulated}). This suggests the presence of a weak evolutionary relationship between ovariole number and egg size, when accounting for body size.

Second, we assessed the relationship between egg size and ovariole number, accounting for body size, within four subclades of insects. We found that across Drosophilidae fly species, egg size is indeed strongly negatively correlated with ovariole number when accounting for body size (Fig. 2c, Table \ref{tab:PGLS-posterior-clade}, p-value `r gsub(0,"<0.001",round(pgls_results_DIP$tTable[[8]],3))`, slope `r round(pgls_results_DIP$coefficients[[2]],3)`, n=`r pgls_results_DIP$dims$N`). For grasshoppers and crickets (Orthoptera), beetles (Coleoptera), and wasps (Hymenoptera), we observed no significant relationship between ovariole number and egg size, even when accounting for body size (Fig. 2d, \ref{fig:PGLS-Coleoptera-Hymenoptera}, Table \ref{tab:PGLS-posterior-clade}, Orthoptera: p-value `r round(pgls_results_ORT$tTable[[8]],3)`, n=`r pgls_results_ORT$dims$N`, Coleoptera: p-value `r round(pgls_results_COL1$tTable[[8]],3)`, n=`r pgls_results_COL1$dims$N`, Hymenoptera: p-value `r round(pgls_results_HYM$tTable[[8]],3)`, n=`r round(pgls_results_HYM$dims$N)`). This indicates that, while a strong negative correlation between egg size and ovariole number exists for some insects, it does not represent a universal pattern across insect clades.

Finally, we tested whether ovariole number is positively correlated with adult body size, and in contrast to previous studies[@honvek1993intraspecific], we found no correlation between ovariole number and adult body mass or length across insects (Fig. \ref{fig:PGLS-body-plot}, Table \ref{tab:PGLS-posterior-body}, body mass: p-value `r round(pgls_results_body_gilbert_name$tTable[[8]],3)`, n=`r pgls_results_body_gilbert_genus$dims$N`, body length: p-value `r round(pgls_results_body_ON_family$tTable[[8]],3)`, n=`r pgls_results_body_ON_family$dims$N`). Of the four subclades considered, only insects in the order Orthoptera had a positive relationship between body size and ovariole number (Table \ref{tab:PGLS-posterior-body}, p-value `r round(pgls_results_body_ORT$tTable[[8]],3)`, slope `r round(pgls_results_body_ORT$coefficients[[2]],3)`, n=`r pgls_results_body_ORT$dims$N`).

![**Tests of the hypothesized trade-off between egg size and ovariole number.** a, Egg volume (mm^3^) and ovariole number, both log~10~ scale; points represent insect species. See section _Modeling ovariole number evolution_ for discussion of the enrichment of certain low values of ovariole number (points appearing vertically arranged) b, Egg volume and ovariole number, residuals to dry adult body mass, points represent genera. c, Drosophilidae egg volume and ovariole number, residuals to thorax length, points represent species. d, Orthoptera egg volume and ovariole number, residuals to body length, points represent genera.](manuscript_figures/Fig_Egg_V2-01.png)

@@

If the hypothesized trade-off between the number and size of offspring is true for insects, then one explanation for the lack of a consistent negative relationship between ovariole number and egg size is that ovariole number may not be a reasonable proxy for offspring number. Previous research has shown that, across individuals within the same species, ovariole number is correlated with certain measurements of fecundity, such as maximum daily rate of egg production for _Drosophila_ [@david1970nombre;@bouletreau1982reproductive], but not others, such as lifetime fecundity[@schmidt2005geographic] or fitness in competition assays[@wayne1997quantitative]. Few studies have compared fecundity and ovariole number across species[@stewart1991reproductive], likely due to the difficulties of measuring fecundity consistently across insects, many of which lay eggs singly and continuously rather than in distinct clutches.

Using a previously reported dataset of lifetime fecundity measurements across insects[@Gilbert2007;@gilbert2010parental], we assessed the relationship between lifetime fecundity and ovariole number. We observed a significant positive relationship (Fig. 3, p-value `r round(pgls_results_fecundity_genus$tTable[[8]],3)`, slope `r round(pgls_results_fecundity_genus$coefficients[[2]],3)`, n=`r pgls_results_fecundity_genus$dims$N`), however, a substantial fraction of repeat analyses show these results are not robust to uncertainty (`r sum(sapply(allometry_fecundity_genus,function(x){x$tTable[[8]]}) > 0.01)` of 1000 regressions are not significant, Table \ref{tab:PGLS-posterior-fecundity}). We note that this relationship is largely defined by the absence of insects with high ovariole number and low fecundity (Fig. 3, empty bottom right corner), while for insects with low ovariole number, fecundity varied over more than three orders of magnitude. We interpret our results, in conjunction with those previously reported, to suggest that ovariole number, when variable across insects in a lineage, may be one factor among many influencing the number of eggs produced. However, we caution against using ovariole number as a direct mathematical proxy for offspring number.

@@

Using the dataset compiled here and a previously published phylogeny of insects (Fig. 5a)[@church2019insect], we modeled the rate of evolutionary change in ovariole number (Figs. \ref{fig:arbutus}, \ref{fig:BAMM-analysis-convergence}, \ref{fig:BAMM-analysis-burnin}, \ref{fig:BAMM-analysis-results}). We observed substantial rate heterogeneity in the evolution of ovariole number (Fig. \ref{fig:BAMM-analysis-results}), meaning that for some lineages ovariole number has evolved rapidly where in others, ovariole number has evolved very slowly or not at all. The most striking example of this are the multiple lineages which have independently evolved invariant or near-invariant ovariole number across taxa (e.g. nearly all Lepidoptera have exactly eight ovarioles, Fig. 5b, \textcolor{red}{Lepidoptera are part of Amphiesmenoptera}, in cyan), from an ancestral variable state. These invariant lineages were identified by finding regions of the phylogeny that experience extremely low rates of ovariole number diversification (Figs. \ref{fig:BAMM-analysis-results}, \ref{fig:BAMM-anlaysis-mean-rates}). Using this approach, we found that invariant ovariole numbers have evolved at least nine times independently across insects, with several subsequent reversals from invariant to variable states (Fig. 5a).

@@

The distribution of ovariole numbers across insects is enriched for even numbers of total ovarioles (Fig. 5c). While many insects show asymmetries in the number of ovarioles between the left and right ovaries, all of the invariant lineages are symmetric (at 4, 6, 8, 10, 12, and 14 total ovarioles). \textcolor{red}{Additionally, for the insects identified as part of invariant lineages, none have any reported intraspecific variation in ovariole number.} Therefore, invariant lineages have near-zero variation when comparing between species, between individuals within a species, and between the left and right ovary within an individual.
